MSstatsPTM: Statistical relative quantification of post-translational modifications in bottom-up mass spectrometry-based proteomics

### 2022-06-13 main/MSstatsPTM_Main.tex
Suggestions for methods

\subsection*{Existing statistical methods for experiments targeting PTMs}

\noindent {\bf Two-sample $t$-test} Many investigations perform differential abundance analysis of PTMs using two-sample $t$-test or its extensions. While simple, these approaches do not fully account for all sources of variations, and are not directly applicable to experiments with complex designs. These include experiments with comparisons of multiple conditions, acquisition in multiple batches, and unbalanced designs. Additionally, while these approaches can be applied to experiments studying PTMs, there is not a self contained, straightforward implementation of the methods, making application challenging.

When modeling the simple experiment seen in Figure \ref{fig:data-structure}, the same Null Hypothesis is addressed as seen in Equation \ref{eq:null_hyp}. To estimate the parameters of interest in the trivial model, the approach takes as input intensities of individual features from modified peptides, or intensity ratios of modified and unmodified peptide features, and compares the mean abundance of a PTM site from one condition to another~\cite{Schwammle2015}\cite{THOMAS2020}. The abundance in each run is taken as input and is often estimated by sum of peak intensities. For example, the log-abundance estimate for the PTM in Run $j$ of Condition $i$ is given by Equation \ref{eq:feature_estimation}.

@@

$t$-test then takes as input the adjusted PTM abundance of each condition and biological replicate, $\hat{u}^{\ast}_{11}$, $\hat{u}^{\ast}_{12}$, $\hat{u}^{\ast}_{21}$, and $\hat{u}^{\ast}_{22}$, to estimate the quantity of interest $\hat{\Delta}$ as well as the standard error, $\widehat{SE(\hat{\Delta})}$. $\hat{\Delta}$ is the estimate of the difference for the Null and Alternative Hypotheses show in Equation \ref{eq:null_hyp}. The statistical significance is determined by comparing the test statistic against the $t$ distribution, with degrees of freedom $df=2J-2$ in balanced designs.

\medskip \noindent {\bf Limma} Modifications of $t$-test such as moderated $t$-test with Limma were also proposed~\cite{Ritchie_15a}\cite{Zhu}. The Null Hypothesis we are interested in answering is the same as in Equation \ref{eq:null_hyp}. To estimate $\hat{\Delta}$ in Figure \ref{fig:data-structure}, feature level summarization and unmodified protein adjustment are performed in the same way as $t$-test, shown in Equations \ref{eq:feature_estimation} and \ref{eq:adj_estimation}. Limma then fits a linear model using the same adjusted summarized features as in $t$-test, $\hat{u}^{\ast}_{11}$, $\hat{u}^{\ast}_{12}$, $\hat{u}^{\ast}_{21}$ and $\hat{u}^{\ast}_{22}$, to estimate $\hat{\Delta}$. To estimate $\widehat{SE(\hat{\Delta})}$, Limma leverages Empirical Bayes moderation to moderate the model variance of the individual PTM using the shared global variance of all PTMs modeled in the experiment. For a single PTM $m$ with an estimated fold change $\hat{\Delta}_m$, the estimated variance is calculated as $s^{adj}_{m} = P(s_{m}| s_\Theta)$ where $s_{m}$ is the variance resulting from the linear model of PTM $m$ and $s_\Theta$ is the global variance across all PTMs. The statistical significance is determined using $\hat{\Delta}_m$ and $\widehat{SE(\hat{\Delta})}$ calculated using the adjusted variance $s^{adj}_{m}$.

@@

The proposed approach leverages the large amount of work done to create $MSstats$~\cite{Choi:2014}. $MSstats$ was designed for protein significance analysis and statistical relative quantification of proteins and peptides in global, targeted and data-independent proteomics. In terms of Figure \ref{fig:data-structure}, $MSstats$ was designed to estimate the difference in the unmodified protein in condition $1$ and condition $2$, denoted by $\mu_1$ and $\mu_2$ respectively. In other words it allows us to answer the new Null Hypothesis seen in Equation \ref{eq:msstats_null_hyp}.

@@

\begin{figure}[ht]
\centering
\includegraphics[scale=.6]{images/fig3.png}
\caption{
\todo{In 2a, it may be more meaningful to replace 'run' with 'bio rep'} Goals of PTM characterization, input to statistical analyses, and notation. (a) Schematic data representation, in a simplified case of two conditions and two biological replicates. Each PTM site is modeled and characterized separately, where a PTM is quantified with observed spectral features (boxes), distinguished by different charge states of a peptide. The feature intensities of the PTM in Condition 1 are denoted by $y_1^\ast$. Features corresponding to unmodified peptides in Condition 1 are denoted by $y_1$. Peptides can be fully cleaved (solid lines) and/or partially cleaved (dashed lines). (b) PTM relative quantification of the population parameters by statistical inference, which makes use of the feature intensities to infer the underlying PTM abundance and protein abundance with an estimate of associated uncertainty. (c) Model-based testing for differential PTM abundance, which corrects for the underlying protein abundance with a cost of increased uncertainty about the estimate of difference between conditions.}
\label{fig:data-structure}
\end{figure}

>>>>

\subsection*{Existing statistical methods for detecting differentially abundant PTMs}

\noindent {\bf Two-sample $t$-test} 
\todo{OV: could you help me with this part a little bit? I tried to organize the text, but could not quite finish. We should have the following order: (1) input, (2) model for the input, (3) model-based inference, (4) implementation; (5) good things and limitations. I think it may help to separate two versions: based on the features, and based on the ratios. Otherwise it is not very clear}
Many investigations perform differential abundance analysis of PTMs using two-sample $t$-test or its extensions. $t$-test takes as input intensities of individual features from modified peptides, or intensity ratios of modified and unmodified peptide features \todo{OV: let's separate the intensities and the ratios}, and compares the mean abundance of a PTM site from one condition to another~\cite{Schwammle2015}\cite{THOMAS2020}. The abundance in each run is taken as input and is often estimated by sum of peak intensities. For example, the log-abundance estimate for the PTM in Run $j$ of Condition $i$ is given by Equation \ref{eq:feature_estimation}. as input the adjusted PTM abundance of each condition and biological replicate, and estimates $\Delta$ in Equation \ref{eq:null_hyp}.
\todo{OV: what is the model?}
To estimate the parameters of interest in this trivial model, the approach takes as input

@@

to estimate the quantity of interest $\hat{\Delta}$ as well as its standard error, $\widehat{SE(\hat{\Delta})}$.  The statistical significance is determined by comparing the test statistic against the $t$ distribution, with degrees of freedom $df=2J-2$ in balanced designs.

While simple, these approaches do not fully account for all sources of variations, and are not directly applicable to experiments with complex designs. These include experiments with comparisons of multiple conditions, acquisition in multiple batches, and unbalanced designs. Additionally, while these approaches can be applied to experiments studying PTMs, there is not a self contained, straightforward implementation of the methods, making application challenging.

\medskip \noindent {\bf Limma} Modifications of $t$-test such as moderated $t$-test with Limma were also proposed~\cite{Ritchie_15a}\cite{Zhu}.

\todo{OV: could you also re-organize: (1) input, (2) model for the input, (3) model-based inference, (4) implementation, (5) good things and limitations. Limitation: limited in experimental design (at most two variance components, with strong assumptions)}

To estimate $\hat{\Delta}$ in Figure \ref{fig:data-structure}, feature level summarization and unmodified protein adjustment are performed in the same way as $t$-test, shown in Equations \ref{eq:feature_estimation} and \ref{eq:adj_estimation}. Limma then fits a linear model using the same adjusted summarized features as in $t$-test, $\hat{u}^{\ast}_{11}$, $\hat{u}^{\ast}_{12}$, $\hat{u}^{\ast}_{21}$ and $\hat{u}^{\ast}_{22}$, to estimate $\hat{\Delta}$. To estimate $\widehat{SE(\hat{\Delta})}$, Limma leverages Empirical Bayes moderation to moderate the model variance of the individual PTM using the shared global variance of all PTMs modeled in the experiment. For a single PTM $m$ with an estimated fold change $\hat{\Delta}_m$, the estimated variance is calculated as $s^{adj}_{m} = P(s_{m}| s_\Theta)$ where $s_{m}$ is the variance resulting from the linear model of PTM $m$ and $s_\Theta$ is the global variance across all PTMs. The statistical significance is determined using $\hat{\Delta}_m$ and $\widehat{SE(\hat{\Delta})}$ calculated using the adjusted variance $s^{adj}_{m}$.

@@

\todo{OV: could you also re-organize: (1) input, (2) model for the input, (3) model-based inference, (4) implementation, (5) good things and limitations. Here good things are: complex designs, including multiple variance components; EB (for TMT); good summarization; implementation. Limitation: MSstats is not designed for protein-level adjustment of changes in PTM}

$MSstats$~\cite{Choi:2014} is designed for protein significance analysis and statistical relative quantification of proteins and peptides in global, targeted and data-independent proteomics. In terms of Figure \ref{fig:data-structure}, $MSstats$ was designed to estimate the difference in the unmodified protein in condition $1$ and condition $2$, denoted by $\mu_1$ and $\mu_2$ respectively. In other words it allows us to answer the new Null Hypothesis seen in Equation \ref{eq:msstats_null_hyp}.

@@

\begin{figure}[ht]
\centering
\includegraphics[scale=.6]{images/fig3.png}
\caption{
\todo{In 2a, it may be more meaningful to replace 'run' with 'bio rep'} Goals of PTM characterization, input to statistical analyses, and notation. (a) Schematic data representation, in a simplified case of two conditions and two biological replicates. Each PTM site is modeled and characterized separately, where a PTM is quantified with observed spectral features (boxes), distinguished by different charge states of a peptide. The feature \todo{log-}intensities of the PTM in Condition 1 are denoted by $y_1^\ast$ \todo{In the figure $y^*$ has 3 subscripts. Could you edit here as well?}. \todo{Log-intensities of } features corresponding to unmodified peptides in Condition 1 are denoted by $y_1$ \todo{In the figure $y$ has 3 subscripts. Could you edit here as well?}. Peptides can be fully cleaved (solid lines) and/or partially cleaved (dashed lines). \todo{OV: I am wondering whether we need Fig 2(b). The null hypothesis is specified in the text. I am not sure it is referenced anywhere. Maybe remove it for simplicity?} (b) PTM relative quantification of the population parameters by statistical inference, which makes use of the feature intensities to infer the underlying PTM abundance and protein abundance with an estimate of associated uncertainty. (c) Model-based testing for differential PTM abundance, which corrects for the underlying protein abundance with a cost of increased uncertainty about the estimate of difference between conditions.}
\label{fig:data-structure}
\end{figure}
